Scaffolder - Software for Reproducible Genome Scaffolding.

### 2011-03-07 src/article.tex
Revise background

High-throughput sequencing can produce hundreds of thousands to millions of
sequence reads from a genome. Each sequence represents the machine-read
nucleotides from a short region in the genome. At the time of writing,
high-throughput sequencing is limited to producing reads less than 1,000
nucleotides in length. Therefore to resolve a sequence longer than this, such
as a complete genome, these numerous smaller fragments must be pieced together.
The process of joining reads into longer sequences is the `assembly' stage of
a genome project \cite{miller2010}. \pb

Assembly software takes the nucleotide reads produced by the sequencing
hardware and, in the ideal case, outputs a complete genome sequence composed of
these individual fragments. An analogy for this process is a jigsaw puzzle:
each nucleotide read represents a single piece, and the final genome sequence
is the completed puzzle. Sequences of repetitive nucleotide `repeat' regions
within the genome, or biased and incomplete sequencing data, may prevent the
genome being assembled into a continuous sequence. This may be due to not
enough, or multiple different overlaps between reads and is analogous to
missing pieces in the jigsaw or pieces that fit to multiple other pieces. \pb

The recent advent of these current high throughput sequencing methods has led
to a renewed interest into algorithms for solving the problem of genome
assembly \cite{pop2008,pop2009}; genome assembly software may however be unable
to produce a complete sequence from these reads. In many cases the algorithm
will generate several large assembled regions of sequence (`contigs') composed
of the many available individual reads, which represent the assembly of reads
into the longest possible sequences given the data. Nonetheless, these contigs
represent a fragmented picture of the genome and require additional work to
join together into larger lengths. \pb

The process of ameliorating a draft sequence into a finished continuous
sequence can be expensive in both time and laboratory effort, although the
genomic data present in a set of contigs may often be sufficient for many
research questions \cite{branscomb2002}. Nevertheless, a continuous
high-quality `finished' genome sequence does provide a greater depth of
information, such as complete resolution of repeat regions and precise
estimates of distances between genomic elements \cite{parkhill2002,fraser2002}.
This process of joining contigs together to form a continuous sequence is
called the `scaffolding' or `finishing' stage. \pb

@@

Scaffolding is the process of joining a series of disconnected contigs into
a complete continuous genome sequence. Due to genomic complexity and missing
data, scaffolding may not ultimately produce a final completed sequence, but
may still succeed in joining a greater subset of contigs together or resolving
gaps. An overview of the required steps in the scaffolding process is outlined
below.

@@

The sequencing process generates reads from either strand of the DNA helix the
resulting contigs constructed from these reads may point in either direction of
the DNA. Orientating contigs to point in the same direction and represent the
same strand requires reverse complementing sequences where necessary. In the
case of microbial genomes this orientation may be to the 5' $\rightarrow$ 3'
direction leading away from the origin of replication.

@@

Given the correct order and orientation, determining the distance between
contigs results in an estimate of the complete genome size. The size of any
inter-contig gaps represents the length of an unknown region in the genome.
Filling these regions with unknown nucleotide characters (`N') allows a draft
continuous sequence, which is useful for representing both the sequenced and
still to be resolved areas in the genome sequence.

@@

During the scaffolding process, closing and filling gaps between contigs
improves the completeness of the genome scaffold. Closing gaps may require
returning to the laboratory to perform additional sequencing or using
computational methods of estimating the unknown sequence. This additional
sequence is then used to replace the gap between two contigs joining them into
a single sequence if they are not already scaffolded together. Once all contigs
have been joined and gaps in a scaffold closed, the genome may be considered
finished.

@@

The process of resolving a genome sequence and building a scaffold uses
\emph{in vitro} methods, \emph{in silico} methods, or a combination of both. An
example of a computational method might use paired-read data, where the
sequencing reads are produced in pairs at a known distance apart in the genome.
The occurrence of a pair of reads in separate contigs can be used to estimate
the order and distance between the contigs. Alternatively, laboratory methods
may use PCR to amplify the unknown DNA in a gap region, then use traditional
Sanger sequencing to determine the gap sequence. Computational methods are more
preferable, as they are less costly in laboratory time and materials compared
to manual gap resolution and uses the available data that exists following
sequencing \cite{nagarajan2010}. \pb

Synteny methods compare the assembled contigs to a more complete reference
genome sequence searching for areas of sequence similarity between the two.
Areas of corresponding sequence between the two sets can be used to infer
contig placement and build the contigs into a scaffold
\cite{richter2007,zhao2008,assefa2009}. Recombination between the two genomes
can, however, reduce the effectiveness of this analysis. Repeat regions may
also be responsible for many of the gaps in building a genome sequence. The
existence of tandemly repeated nucleotide regions produces many similar reads
fromgenome sequencing. The assembly of reads generated from repeat regions can
result in the assembly collapsing into a single repeat instance or ignored by
more conservative assembly algorithms. These regions may however be resolved by
using reassembly algorithms specifically for repeat resolution
\cite{mulyukov2002,koren2010}. Similarly a related approach may use unassembled
sequence reads matching the regions around a scaffold gap to construct
a uniquely overlapping set of reads across the gap \cite{tsai2010}. \pb

Paired-read data provides an extra level of information about how contigs may
be scaffolded together. Heuristic algorithms may take advantage of this data to
search for optimal configuration of contigs in the scaffold using paired-read
distances \cite{dayarian2010,boetzer2011}. Additional data using a reference
genome may also be combined when estimating the best contig configuration
\cite{pop2004}. Finally when the scaffold cannot be completely resolved,
\emph{in silico} software packages exist to suggest the likely primers
necessary for PCR amplifying the sequence in gap regions
\cite{gordon2001,nagarajan2010}. \pb

These methods provide a wide array of approaches for merging contigs into
larger continuous scaffold sequences. The scaffolding process may still however
require inserting additional sequence or joining contigs using PCR-derived
information from the laboratory. This manual stage in the finishing process
comprises moving and editing large blocks of nucleotide text by hand which is
subject to human error and precludes any reproducibility of the steps taken.
\pb

The software outlined here, ``Scaffolder'', aims to address these problems by
creating a file syntax and software framework for editing a genome scaffold.
Scaffolder uses a specific file format to define how contigs are joined,
additional sequences inserted, and for the specification of unknown regions.
This syntax allows a scaffold to be updated by simply editing the scaffold
file. Scaffolder facilites a reproducible scaffolding process and provides
a concise overview of how the scaffold was constructed. \pb

>>>>

High-throughput sequencing can produce hundreds of thousands to millions of
sequence reads from a genome. At the time of writing, high-throughput
sequencing is limited to producing reads less than 1,000 nucleotides in length.
Therefore to resolve a sequence longer than this, such as a complete genome,
these numerous smaller fragments must be pieced together. The process of
joining reads into longer sequences is the `assembly' stage of a genome project
\cite{miller2010}. \pb

Assembly software takes the nucleotide reads produced by sequencing hardware
and, in the ideal case, outputs a single complete genome sequence composed of
these individual fragments. An analogy for this process is a jigsaw puzzle:
each nucleotide read represents a single piece, and the final genome sequence
is the completed puzzle. Sequences of repetitive nucleotide `repeat' regions
within the genome, or biased and incomplete sequencing data, may prevent the
genome being assembled into a continuous sequence. This may be due to not
enough, or multiple different overlaps between reads and is analogous to
missing pieces in the jigsaw or pieces that fit to multiple other pieces. \pb

The advent of high throughput sequencing methods has led to a renewed interest
in algorithms to solve the problem of genome assembly \cite{pop2008,pop2009}.
Nonetheless, due to the complexity of merging large numbers of reads genome
assembly software may be unable to produce a complete sequence from these
reads. Instead the algorithm may generate several large assembled regions of
sequence (`contigs') composed of the many available individual reads that
represent the assembly of reads into the longest possible sequences, given the
data. Nonetheless, these contigs represent a fragmented picture of the genome
and require additional work to join together into a complete sequence. \pb

The process of ameliorating a draft sequence into a finished continuous
sequence can be expensive in terms of time and laboratory effort, although the
genomic data present in a set of generated contigs may be sufficient for many
research questions \cite{branscomb2002}. Nevertheless, a continuous
high-quality `finished' genome sequence does provide a greater depth of
information, such as complete resolution of repeat regions and precise
estimates of distances between genomic elements \cite{parkhill2002,fraser2002}.
The process of joining these contigs together to form a continuous genome
sequence is called the `scaffolding' or `finishing' stage. \pb

@@

Scaffolding is the process of joining a series of disconnected contigs into
a complete continuous genome sequence. Due to genomic complexity and missing
data, scaffolding may not ultimately produce a final completed sequence, but
may still succeed in a greater subset of contigs together or resolving gaps. An
overview of the required steps in the scaffolding process is outlined below:

@@

The sequencing process generates reads from either strand of the DNA helix and
the resulting contigs constructed from these reads may represent either the
5'~$\rightarrow$~3' or 3'~$\rightarrow$~5' orientation with respect to the
origin of replication. Orientating all contigs to point in the same direction
requires reverse complementing sequences where necessary. In the case of
microbial genomes this orientation will be to the 5'~$\rightarrow$~3' direction
leading away from the origin of replication.

@@

Given the correct order and orientation, determining the distance between
contigs results in an estimate of the complete genome size. The size of any
inter-contig gaps represents the length of an unknown region in the genome.
Filling these regions with unknown nucleotide characters (`N') allows a draft
continuous sequence, which is useful for representing both the sequenced and
to-be-resolved areas in the genome sequence.

@@

During the scaffolding process, closing and filling gaps between contigs
completes and improves the genome scaffold. Closing gaps may require returning
to the laboratory to perform additional sequencing or using computational
methods of estimating the unknown sequence. This additional sequence is then
used to replace the gap between two contigs, joining them into a single
sequence. Once all contigs have been joined and gaps in a scaffold closed, the
genome may be considered finished.

@@

The process of resolving a genome sequence and building a scaffold uses wet
laboratory methods, \emph{in silico} methods, or a combination of both. An
example of a computational method might use the existing paired-read data from
the sequencing stage where reads were produced in pairs at a known distance
apart in the genome. The occurrence of paired reads in separate contigs can be
used to probabalistically estimate the order and distance between these
contigs. Alternatively, laboratory methods may use PCR to amplify the unknown
DNA in a gap region, then use traditional Sanger sequencing to determine the
sequence of this gap. Computational methods are more preferable as they are
less costly in laboratory time and materials compared to manual gap resolution,
and use the available data that exists following sequencing
\cite{nagarajan2010}. Finally when the scaffold cannot be completely resolved,
\emph{in silico} software packages exist to suggest the likely primers
necessary for PCR amplifying the sequence in gap regions \cite{gordon2001}. \pb

Synteny methods for scaffolding compare the assembled contigs to a more
complete reference genome sequence by searching for areas of sequence
similarity between the two. Any areas of corresponding sequence with the
reference genome can be used to infer contig placement and build the contigs
into a scaffold \cite{richter2007,zhao2008,assefa2009}. Recombination within
each of the two compared genomes can, however, reduce the effectiveness of this
analysis. \pb

Repeat regions may also be responsible for multiple gaps when building a genome
sequence; tandemly repeated nucleotide regions in the genome produces multiple
reads with similar sequence. As many assembly algorithms rely on sequence
overlaps between reads to build a contig the similarity between repeat-region
reads can result in the assembly collapsing into an artifically short sequence
at this region or being ignored by more conservative assembly algorithms. Such
regions can be resolved by using algorithms to specifically reassemble the
collapsed repeat region correctly \cite{mulyukov2002,koren2010}, while
a related approach uses unassembled sequence reads matching the regions around
a scaffold gap to construct a uniquely overlapping set of reads across it
\cite{tsai2010}. \pb

Paired-read data can provide an extra level of information about how contigs
may be scaffolded together. Heuristic scaffolding algorithms take advantage of
this data to search for optimal configuration of contigs in the scaffold using
paired-read distances \cite{dayarian2010,boetzer2011}. Synteny data from
a reference genome can also be combined with this paired-read data to estimate
the best contig configuration \cite{pop2004}. \pb

These described methods provide a wide array of approaches for merging contigs
into a larger, continuous scaffold sequence. Nevertheless the scaffolding
process may still require inserting additional sequences or further joining
contigs using PCR-derived sequence. This manual stage in the finishing
processmay require moving and editing large blocks of nucleotide text by hand
which is subject to human error and precludes any reproducibility of the steps
in this process. \pb

The software outlined here, ``Scaffolder'', aims to address these problems of
reproducibility by creating a file syntax and software framework for editing
a genome scaffold. Scaffolder uses a specific file format to define how contigs
are joined, additional sequences are inserted, and for the specification of
unknown regions. This syntax allows a scaffold to be updated by simply editing
the scaffold file. As such scaffolder facilitates a reproducible scaffolding
process and provides a concise overview of how the final genomic scaffold was
constructed. \pb
